High resolution, annual maps of the characteristics of smallholder-dominated croplands at national scales

### 2021-03-09 inst/manuscript/si.Rmd
Edited ms and si: first upload to EarthArXiv. Edited scripts for data_for_instances.R, added new script to data-raw (aois_tiles.R) to show creation of degree and tiling grids

---
title: "Supporting Information: High resolution, annual maps of the characteristics of smallholder-dominated croplands at national scales"

output:
  pdf_document:
    fig_caption: yes
    fig_width: 7
    fig_height: 7
    keep_tex: yes
    number_sections: yes
    template: manuscript.latex
    includes:
      in_header: header.tex

@@

knitr::opts_chunk$set(fig.pos = "!ht", out.extra = "")

@@

# Methods
## Assessing quality of composites
The rubric presented in Table \ref{tab:imqcriteria} was used to assess the quality of the seasonal image composites. The imagery was evaluated by examining their Raster Foundry overlays within a `labeller` instance set up for the purpose.

@@

### Mapping rules
Labellers were asked to follow a particular set of rules when digitizing field boundaries. The following rules are reproduced (with some modification to wording) from a [website](https://github.com/agroimpacts/mappingrules) that we developed to host the rules.

@@

After collecting the sample, the geometries were further refined by converting the points (sites were the target didn't have to be shifted) to polygons with the same  0.0002866$^\circ$ resolution, and then the full set of map reference polygons was used to extract the classified values from both the cropland probability and vectorized field boundary maps.

\clearpage
# Results
## Image quality
```{r, echo=FALSE, message=FALSE, warning=FALSE}
data("image_quality")
wqual_diff <- image_quality$img %>% filter(kml_type == "F") %>%
  group_by(name, season, worker_id) %>%
  summarize(tile = min(tile), score = sum(score) / 12) %>% 
  ungroup() %>% 
  pivot_wider(names_from = worker_id, values_from = score,
              names_prefix = "w") %>%
  select(tile, season, w92, w99) %>% 
  mutate(d = w92 - w99, 
         absd = abs(d))
```

@@

## Active learning process
### Labelling
The distributions of the initial 500 sites used to train the RandomForest model for each AOI are shown in Figure \ref{fig:trainval}. In AOI 3, the initial active learning cycle resulted in low accuracy because the northern part of the AOI shows low contrast between fields and the surrounding vegetation in the dry season. Training the model with the initial 500 samples resulted in large commission errors in this part of the AOI, thus we ran a second active learning cycle that began with an inital random draw of 300 training sites confined to this AOI (blue points in Figure \ref{fig:trainval}A)
```{r trainval, echo = FALSE, out.width="80%", fig.cap="The distribution of A) initial randomly selected training sites, including 300 points selected within AOI 3 to retrain a second active learning run for this region, as well as the locations of training reference sites, and B) validation points and training sites selected during each active learning iteration. Grey background lines indicate the boundaries of Ghana's agroecozones, with the grey numbers indicating the agroecoregion ID: 1 = Coastal savanna; 2 = Wet evergreen; 3 = Moist evergreen; 4 = Deciduous forest; 5 = Transitional zone; 6 = Guinea savanna; 7 = Sudan savanna; 8 (not shown) = Volta Lake.", fig.align='center', message=FALSE, results='hold'}
knitr::include_graphics('figures/si_training_validation_pts.png')
```

The distribution of data collection effort and scores from training reference sites by labelling team are shown in Figure \ref{fig:trainqual}. These results include those from re-labelling the initial training in Cluster 2 (AOIs 7-9, 12, 15). This was done because we discovered a small spatial offset in the original image composites, which we corrected by reprocessing the images. We replaced the image overlays in the instance for central-southeast \ref{fig:trainval}A), and the labelling team mapped these sites a second time during the production run in late 2019. The reprocessed labels were used to initially train the model for AOIs 9, 12, and 15.

```{r trainqual, echo = FALSE, out.width="60%", fig.cap="The A) distributions of scores at training reference sites for each labeller (means indicated by X in boxplots), and B) the number of training and training reference sites completed by each labeller. Labellers' identities are anonymized.", fig.align='center', message=FALSE, results='hold'}
knitr::include_graphics('figures/si_training_qual_n.png')
```

@@

```{r labelrisk, echo = FALSE, out.width="80%", fig.cap="The average Bayes Risk of each training and validation site in Ghana.", fig.align='center', message=FALSE, results='hold'}
knitr::include_graphics('figures/si_label_risk_map.png')
```

```{r labelriskhist, echo = FALSE, out.width="40%", fig.cap="The distribution of Bayesian Risk values for training (left column) and validation (right column) sites in each AOI, with the average value indicated by vertical lines. The first three rows (panel titles beginning with 'Cl') indicate distributions of Bayes Risk in the initial training clusters, including the first ('Cl2a') and second ('Cl2b') mappings of Cluster 2.", fig.align='center', message=FALSE, results='hold'}
knitr::include_graphics('figures/si_label_risk_hists.png')
```

### Validation
The differences in accuracy, AUC, and F1 between the active learning process and the random retraining at each iteration for AOIs 1, 8, and 15 (Figure \ref{fig:randomvactive}). Small differences due to random variations in the RandomForest models are the reason for the non-zero differences at iteration 0, when both models were trained with the same level set.   
```{r randomvactive, echo = FALSE, out.width="90%", fig.cap="The percent difference in performance metrics per iteration for AOIs 1, 8, and 15 (grey lines with numbers indicating AOI; the black line indicates average difference across the three AOIs) when comparing models trained using active learning versus those trained using randomly selected sites. Postive percentages indicate superior performance by active learning, negative percentages the inverse.", fig.align='center', message=FALSE, results='hold'}
knitr::include_graphics('figures/si_random_vs_active.png')
```

### The impact of training data error
Probability images resulting from Random Forest models trained with labels generated by 1) consensus and the 2) most and 3) least accurate worker to digitize each training site are illustrated in Figure \ref{fig:labelstrategy}. These images were created for a single tile in AOI 1.

```{r labelstrategy, echo = FALSE, out.width="100%", fig.cap="Cropland probability images produced by RandomForest models trained with A) consensus labels, B) the most accurate worker's labels, and C) the least accurate worker's labels.", fig.align='center', message=FALSE}
knitr::include_graphics('figures/si_label_strategy_probs.png')
```

@@

cap <- glue::glue("Spearman's Rank correlation between the average label risk per AOI and a variety of model performance metrics.")
kable(risk_corr, format = "latex", booktabs = "T", caption = cap) %>% 
  kable_styling(full_width = TRUE, latex_options = c("hold_position")) #%>% 
  # column_spec(1, width = "8cm")

# tp <- 10
# fp <- 2
# tn <- 20
# fn <- 3
# tss <- (tp * tn - fp * fn) / ((tp + fn) * (fp + tn))
# mcc <- (tp * tn - fp * fn) / sqrt((tp + fp) * (tp + fn) * (tn + fp) * (tn + fn))

@@

```{r aoizones, echo = FALSE, out.width="100%", fig.cap="Four zones used to combine the A) AOIs and B) agroecozones (B), in order to assess sub-national mapping accuracy. In B, the ID number of Ghana's individual agroecozones is shown in grey: 1 = Coastal savanna; 2 = Wet evergreen; 3 = Moist evergreen; 4 = Deciduous forest; 5 = Transitional zone; 6 = Guinea savanna; 7 = Sudan savanna; 8 (not shown) = Volta Lake.", fig.align='center', message=FALSE, results='hold'}
knitr::include_graphics('figures/si_aoi_zones.png')
```

\begin{table}
\caption{Map accuracies and adjusted area estimates for the ~3 m pixel-wise classifications (based on RandomForest predictions). Results are provided for four different groupings of Ghana's 8 agroecozones zones (Zone 1 = Coastal savanna; Zone 2 = Wet evergreen, Moist evergreen, and Deciduous forest; Zone 3 = Transitional zone; Zone 4 = Guinea savanna and Sudan savanna) plus the entire country. The error matrix (with reference values in columns) provides the areal percentage for each cell, and the producer's (P), user's (U), and overall (O) map accuracies and their margins of error (in parenthesis) are provided, as well as the sample-adjusted area estimates (in km$^{2}$) and margins of error.}
\includegraphics[width = 18cm]{figures/si_aez_accuracies.png}
\label{tab:aezaccuracy}
\end{table}

### Field area and number
To mean, median, and distributions of the average area of segmented field boundaries over the 100 validation sites in each AOI are compared to the areas of the polygons digitized by the most accurate labeller over the same sites in Figure \ref{fig:areavalidation}. The same statistics for average number of segments versus average number of labelled polygons across validation sites in each AOI are shown in Figure \ref{fig:numbervalidation}.      
```{r areavalidation, echo = FALSE, out.width="100%", fig.cap="The distributions of the average areas (in hectares) of segmented field boundaries (shown in blue) at the 100 validation sites per AOI, compared with the average areas of field boundaries digitized (shown in red) by the most accurate worker to label each site. Vertical lines indicate the mean and median of each distribution.", fig.align='center', message=FALSE, results='hold'}
knitr::include_graphics('figures/si_validation_stats_fha2.png')
```

```{r numbervalidation, echo = FALSE, out.width="90%", fig.cap="The distributions of the average number of segmented field boundaries (shown in blue) at the 100 validation sites per AOI, compared with the average number of digitized polygons (shown in red) by the most accurate worker to label each site. The mean and median of each distribution is shown. Vertical lines indicate the mean and median of each distribution.", fig.align='center', message=FALSE, results='hold'}
knitr::include_graphics('figures/si_validation_stats_fn2.png')
```

@@

```{r fieldsizes, echo = FALSE, out.width="100%", fig.cap="The A) average sizes of fields and B) total number of fields in each 0.05 degree tile, as calculated from the vectorized field boundaries.", fig.align='center', message=FALSE, results='hold'}
knitr::include_graphics('figures/si_fsize_n_05.png')
```

>>>>

---
title: "High resolution, annual maps of the characteristics of smallholder-dominated croplands at national scales"

subtitle: "Supporting Information"

output:
  pdf_document:
    fig_caption: yes
    fig_width: 7
    fig_height: 7
    keep_tex: yes
    number_sections: no
    extra_dependencies: ["float", "titling", "setspace"]
    includes:
      in_header: header.tex

@@

knitr::opts_chunk$set(fig.pos = "!ht", out.extra = "")
# knitr::opts_chunk$set(fig.pos = "!H", out.extra = "")

@@

# Methods
## Assessing quality of composites
The rubric presented in Table \ref{tab:imqcriteria} was used to assess the quality of the seasonal image composites. The imagery was evaluated by examining their Raster Foundry [@azaveaRasterFoundry2020] overlays within a `labeller` instance set up for the purpose.

@@

### Mapping rules
Labellers followed set of rules described in the next two sections when interpreting and digitizing field boundaries.

@@

After collecting the sample, the geometries were further refined by converting the points (sites were the target didn't have to be shifted) to polygons with the same  0.0002866$^\circ$ resolution, and then the full set of map reference polygons was used to extract the classified values from both the cropland probability and vectorized field boundary maps. Accuracies were assessed for the entire country, and with several zones consisting of different groupings of AOIs or agroecozones\footnote{sourced from https://wheregeospatial.com/agro-ecological-zones-ghana/}

# Results
## Image catalog and quality

```{r, echo=FALSE, message=FALSE, warning=FALSE}
data("image_quality")
wqual_diff <- image_quality$img %>% filter(kml_type == "F") %>%
  group_by(name, season, worker_id) %>%
  summarize(tile = min(tile), score = sum(score) / 12) %>% 
  ungroup() %>% 
  pivot_wider(names_from = worker_id, values_from = score,
              names_prefix = "w") %>%
  dplyr::select(tile, season, w92, w99) %>% 
  mutate(d = w92 - w99, 
         absd = abs(d))
```

The total number of PlanetScope images available for download from the Planet API [@planetteamPlanetApplicationProgram2018] per tile for each of the two major seasons is shown in Figure \ref{fig:imagedensity}.

```{r imagedensity, echo = FALSE, out.width="90%", fig.cap="The density of images available through the Planet API for the five month growing season in 2018 and the subsequent three month dry season (excluding the month of November), shown in relation to the AOI boundaries.", fig.align='center', message=FALSE, results='hold'}
knitr::include_graphics('figures/si_planet_coverage.png')
```

@@

## Active learning process
### Labelling
The locations of training, training reference, and validation points are shown in Figure \ref{fig:trainval}. In AOI 3, the initial active learning cycle resulted in low accuracy because the northern part of the AOI shows low contrast between fields and the surrounding vegetation in the dry season. Training the model with the initial 500 samples resulted in large commission errors in this part of the AOI, thus we ran a second active learning cycle that began with an inital random draw of 300 training sites confined to this AOI (blue points in Figure \ref{fig:trainval}A)
```{r trainval, echo = FALSE, out.width="70%", fig.cap="The distribution of A) initial randomly selected training sites, including 300 points selected to initialize a second run for retrain AOI 3, and the locations of training reference sites, and B) validation points and training sites selected during each active learning iteration. Grey background lines and show the boundaries and ID number of Ghana's agroecozones: 1 = Coastal savanna; 2 = Wet evergreen; 3 = Moist evergreen; 4 = Deciduous forest; 5 = Transitional zone; 6 = Guinea savanna; 7 = Sudan savanna; 8 (not shown) = Volta Lake.", fig.align='center', message=FALSE}
knitr::include_graphics('figures/si_training_validation_pts.png')
```

The distribution of data collection effort and scores from training reference sites by labelling team are shown in Figure \ref{fig:trainqual}. These results include those from re-labelling the initial training in Cluster 2 (AOIs 7-9, 12, 15). This was done because we discovered a small spatial offset in the original image composites, which we corrected by reprocessing the images. We replaced the image overlays in the instance for central-southeast (Figure \ref{fig:trainval}A), and the labelling team mapped these sites a second time during the production run in late 2019. The reprocessed labels were used to initially train the model for AOIs 9, 12, and 15.

```{r trainqual, echo = FALSE, out.width="50%", fig.cap="The A) distributions of scores at training reference sites for each labeller (means indicated by X in boxplots), and B) the number of training and training reference sites completed by each labeller. Labellers' identities are anonymized.", fig.align='center', message=FALSE}
knitr::include_graphics('figures/si_training_qual_n.png')
```

@@

```{r labelrisk, echo = FALSE, out.width="80%", fig.cap="The average Bayes Risk of each training and validation site in Ghana.", fig.align='center', message=FALSE}
knitr::include_graphics('figures/si_label_risk_map.png')
```

```{r labelriskhist, echo = FALSE, out.width="40%", fig.cap="The distribution of Bayesian Risk values for training (left column) and validation (right column) sites in each AOI, with the average value indicated by vertical lines. The first three rows (panel titles beginning with 'Cl') indicate distributions of Bayes Risk in the initial training clusters, including the first ('Cl2a') and second ('Cl2b') mappings of Cluster 2.", fig.align='center', message=FALSE}
knitr::include_graphics('figures/si_label_risk_hists.png')
```

### Validation
The differences in accuracy, AUC, and F1 between the active learning process and the random retraining at each iteration for AOIs 1, 8, and 15 (Figure \ref{fig:randomvactive}). Small differences due to random variations in the RandomForest models are the reason for the non-zero differences at iteration 0, when both models were trained with the same level set.   
```{r randomvactive, echo = FALSE, out.width="90%", fig.cap="The percent difference in performance metrics per iteration for AOIs 1, 8, and 15 (grey lines with numbers indicating AOI; the black line indicates average difference across the three AOIs) when comparing models trained using active learning versus those trained using randomly selected sites. Postive percentages indicate superior performance by active learning, negative percentages the inverse.", fig.align='center', message=FALSE}
knitr::include_graphics('figures/si_random_vs_active.png')
```

### The impact of training data error
Probability images resulting from Random Forests models trained with labels generated under three different labeling strategies are illustrated in Figure \ref{fig:labelstrategy}. These included consensus labels, and those individual labels that were likely to be the most and least accurate for each training site. Label accuracy was based on the mean score of each labeller against the training reference sites (Figure \ref{fig:trainqual}), as assessed when labelling a given AOI. These images were created for a single tile in AOI 1.

```{r labelstrategy, echo = FALSE, out.width="100%", fig.cap="Cropland probability images produced by Random Forests models trained with A) consensus labels, B) the most accurate individual labels, and C) the least accurate individual labels.", fig.align='center', message=FALSE}
knitr::include_graphics('figures/si_label_strategy_probs.png')
```

@@

cap <- glue::glue("Spearman's Rank correlation between the average label risk per AOI and a variety of model performance metrics.")
kable(risk_corr, format = "latex", booktabs = "T", caption = cap) %>% 
  kable_styling(full_width = TRUE, latex_options = c("hold_position")) #%>% 
  # column_spec(1, width = "8cm")
```

@@

```{r aoizones, echo = FALSE, out.width="90%", fig.cap="Four zones used to combine the A) AOIs and B) agroecozones (B), in order to assess sub-national mapping accuracy. In B, the ID number of Ghana's individual agroecozones is shown in grey: 1 = Coastal savanna; 2 = Wet evergreen; 3 = Moist evergreen; 4 = Deciduous forest; 5 = Transitional zone; 6 = Guinea savanna; 7 = Sudan savanna; 8 (not shown) = Volta Lake.", fig.align='center', message=FALSE, results='hold'}
knitr::include_graphics('figures/si_aoi_zones.png')
```

\begin{table}
\caption{Map accuracies and adjusted area estimates for the ~3 m pixel-wise classifications (based on RandomForest predictions). Results are provided for four different groupings of Ghana's 8 agroecozones zones (Zone 1 = Coastal savanna; Zone 2 = Wet evergreen, Moist evergreen, and Deciduous forest; Zone 3 = Transitional zone; Zone 4 = Guinea savanna and Sudan savanna) plus the entire country. The error matrix (with reference values in columns) provides the areal percentage for each cell, and the producer's (P), user's (U), and overall (O) map accuracies and their margins of error (in parenthesis) are provided, as well as the sample-adjusted area estimates (in km$^{2}$) and margins of error.}
\begin{center}
\includegraphics[width = 16cm]{figures/si_aez_accuracies.png}
\end{center}
\label{tab:aezaccuracy}
\end{table}

### Field area and number
To mean, median, and distributions of the average area of segmented field boundaries over the 100 validation sites in each AOI are compared to the areas of the polygons digitized by the most accurate labeller over the same sites in Figure \ref{fig:areavalidation}. The same statistics for average number of segments versus average number of labelled polygons across validation sites in each AOI are shown in Figure \ref{fig:numbervalidation}.      
```{r areavalidation, echo = FALSE, out.width="90%", fig.cap="The distributions of the average areas (in hectares) of segmented field boundaries (shown in blue) at the 100 validation sites per AOI, compared with the average areas of field boundaries digitized (shown in red) by the most accurate worker to label each site. Vertical lines indicate the mean and median of each distribution.", fig.align='center', message=FALSE}
knitr::include_graphics('figures/si_validation_stats_fha2.png')
```

```{r numbervalidation, echo = FALSE, out.width="90%", fig.cap="The distributions of the average number of segmented field boundaries (shown in blue) at the 100 validation sites per AOI, compared with the average number of digitized polygons (shown in red) by the most accurate worker to label each site. The mean and median of each distribution is shown. Vertical lines indicate the mean and median of each distribution.", fig.align='center', message=FALSE}
knitr::include_graphics('figures/si_validation_stats_fn2.png')
```

@@

```{r fieldsizes, echo = FALSE, out.width="90%", fig.cap="The A) average sizes of fields and B) total number of fields in each 0.05 degree tile, as calculated from the vectorized field boundaries.", fig.align='center', message=FALSE}
knitr::include_graphics('figures/si_fsize_n_05.png')
```
